---
title: "R Club - *decontam* solutions"
date: "4/4/2024"
author: "Yasmin Hilliam, PhD"
output: html_notebook
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```
### Package set up
Install and load required packages
```{r package loading, results = FALSE, message = FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install()
```
```{r, results = FALSE, message = FALSE}
BiocManager::install(pkgs = c("tidyverse","phyloseq","decontam"))
```
```{r, message = FALSE}
library(tidyverse)
library(phyloseq) 
library(decontam)
```
### Plot theme set up
Sets up a global theme for all plots created during this session
```{r, message = FALSE}
theme_set(theme_bw()) # set global theme to bw
theme_replace(axis.title.x=element_text(size=14), # set individual theme options for plotting
              axis.title.y=element_text(size=14,angle=90),
              axis.text.x=element_text(size=12),
              axis.text.y=element_text(size=12),
              legend.title=element_text(size=14),
              legend.text=element_text(size=12))
```
## Activities
### Combined models
Prevalence and frequency are the two basic models by which *decontam* identifies contaminants. However, the package also offers models that combine both methods.

1. Run the help command to open the *decontam* manual and read about the other available methods.
```{r, message = FALSE}
?isContaminant
```

2. Identify contaminants using the "both" method. How many contaminants are identified using this method?
```{r, message = FALSE}
contamdfBoth <- isContaminant(ps,
                              method = "both",
                              neg = "is.neg",
                              conc = "quant")

table(contamdfBoth$contaminant)
```

3. Generate a plot of prevalence in controls vs. samples. Does the "both" method seem suitable for this dataset? Why?
```{r, message = FALSE}
dfBoth <- data.frame(prevPos = taxa_sums(psPrev.pos), # build dataframe
                     prevNeg = taxa_sums(psPrev.neg),
                     contaminant = contamdfBoth$contaminant) # change source of contaminant identification to combined dataset

ggplot(data = dfBoth, # input dataframe
       aes(x = prevNeg, # x axis prevalence in negative controls
           y = prevPos, # y axis prevalence in samples
           color = contaminant)) + 
  geom_point(shape = 1) +
  xlab("Prevalence (controls)") + 
  ylab("Prevalence (samples)")
```

When using methods that combine both frequency and prevalence calculations, you can provide a vector with a length of two to the threshold argument: the first value sets the threshold for the frequency test and the second value sets the threshold for the prevalence test. Using the "either" method, attempt to optimise these values. Plot prevalence in samples vs. controls to check that ASVs of concern are being correctly identified as contaminants.

4. Identify contaminants using "either" method
```{r, message = FALSE}
contamdfEith <- isContaminant(ps,
                              method = "either",
                              neg = "is.neg",
                              conc = "quant",
                              threshold = c(0.1, 0.5))

table(contamdfEith$contaminant)
```

5. Plot prevalence in samples vs. controls
```{r, message = FALSE}
dfEith <- data.frame(prevPos = taxa_sums(psPrev.pos), # build dataframe
                     prevNeg = taxa_sums(psPrev.neg),
                     contaminant = contamdfEith$contaminant)

ggplot(data = dfEith, # input dataframe
       aes(x = prevNeg, # x axis prevalence in negative controls
           y = prevPos, # y axis prevalence in samples
           color = contaminant)) + 
  geom_point(shape = 1) +
  xlab("Prevalence (controls)") + 
  ylab("Prevalence (samples)")
```